---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This report demonstrates some of the analysis of daily activities that is 
possible using personal activity monitoring devices (like a Fitbit, Nike 
Fuelband, or Jawbone Up) and a powerful data analysis language like R.
Specifically, this report will illustrate the R code necessary to input, 
process, visualize, and analyze the activity data collected by an anonymous
individual in the months of October and November 2012.

## Loading and preprocessing the data

The data for this analysis has been archived as a zip file, so the first step
will be to extract the files from the archive and save them in a data directory, 
but only if the files haven't been extracted previously. The following R code
will check to see if the data directory has already been created and will
unzip the archive if it hasn't.
```{r}
if(!file.exists("./data")) {    
    unzip("activity.zip", exdir = "./data", unzip = "internal")
}
```

Now that the data has been extracted from the archive, the contents of the data
directory can be listed.
```{r}
list.files("./data")
```

The data file contains a single comma-separated values (CSV) file, so the next
step is to read the file contents into an R object called activityData so it can
be analyzed.
```{r}
activityData <- read.csv("./data/activity.csv")
```

Viewing the first 15 rows of the R object and a summary will show what the
variable names are, what the data values look like, and what their ranges are.
```{r}
head(activityData, 15)
summary(activityData)
```

It appears that the data file contains a *steps* variable, which may not be
available, the *date* of the observation in Year-Month-Day format, and an 
*interval* between observations of five minutes with the hundreds positions
indicating the hours (e.g., 105 indicates the interval at 1:05). The summary
also shows that there appear to be 288 observations per day and the maximum 
range of the *interval* is from 0 to 2355 (24 hours of observations).

It may be easier for later analysis to convert the *date* and *interval*
variables into a standard R date and time variable. The following R code uses
the **formatC** function to convert the varying length *interval* values to four
digits with leading zeros and the **strptime** function to convert the *date* 
and *interval* values into POSIX data-time format and adds them as a new column 
in *activityData*.
```{r}
activityData$datetime <- strptime(paste(activityData$date,
                                        formatC(activityData$interval,
                                                width = 4,
                                                format = "d",
                                                flag = "0")),
                                        format = "%Y-%m-%d %H%M")
head(activityData)
```


## What is mean total number of steps taken per day?

The total number of steps per day can be visualized as a histogram using the
following code.
```{r}
stepsPerDay <- aggregate(activityData$steps ~ activityData$date, 
                         activityData, 
                         sum)
hist(stepsPerDay[ , 2], 
     main = "Total Steps Taken Per Day", 
     xlab = "Total Steps")
```

The mean of the total number of steps per day can be calculated as follows.
```{r}
mean(stepsPerDay[ , 2])
```

Likewise, the median can also be determined.
```{r}
median(stepsPerDay[ , 2])
```

## What is the average daily activity pattern?

The following R code calculates and plots the average number of steps per time
interval.
```{r}
stepsPerInterval <- aggregate(activityData$steps ~ activityData$interval,
                              activityData, 
                              ave)
plot(stepsPerInterval[, 1], stepsPerInterval[ , 2][ , 1], type = "l", 
     main = "Average Steps Per Interval",
     xlab = "Interval", ylab = "Average Steps")
```

The following code displays the interval for which the maximum average steps 
per interval was calculated.
```{r}
stepsPerInterval[max(stepsPerInterval[ , 2][ , 1]), 1]
```

## Imputing missing values

The following R code shows how many missing ("NA") values are in the data set
for the *steps* data.
```{r}
sum(is.na(activityData$steps) == TRUE)
```

There are a number of possible strategies for replacing the missing values in
the data set. In the last section the average number of steps per time
interval over all dates was calculated. The average could be used to replace the
"NA" value in the missing time intervals. In the section prior to the last one, 
the daily average steps were calculated. The daily average could also be
substituted for the missing values. However, in this dataset, the activity of
the wearer of the device is being tracked. Assuming this is for the purpose of
improving fitness, adding average values could very well overestimate the 
activity of the wearer, so I have decided to just replace all of the "NA"
values with zeroes. This code creates a new data frame with all numeric values
(no "NAs").
```{r}
numericActivityData <- activityData
numericActivityData[is.na(numericActivityData)] <- 0
head(numericActivityData)
```

Comparing the contents of this data set with the following listing of the first
six rows of the original data set demonstrates that the NAs have been replaced
with zeroes.
```{r}
head(activityData)
```

The following code displays the histogram of the total number of steps per day
for this new data set.
```{r}
numericStepsPerDay <- aggregate(numericActivityData$steps ~ 
                                    numericActivityData$date,
                                numericActivityData,
                                sum)
hist(numericStepsPerDay[ , 2], 
     main = "Total Steps Taken Per Day (No NAs)", 
     xlab = "Total Steps")
```

Comparing this histogram to the one generated previously, it is apparent that 
ignoring the NAs rather than replacing them with numeric values did have an 
effect on the results. The following is the mean for the data set
with all numeric values.

```{r}
mean(numericStepsPerDay[ , 2])
```

And the following code calculates the median for this new data set.
```{r}
median(numericStepsPerDay[ , 2])
```

The difference between the original data set and this data set with all 
numeric values can be quantified by taking the difference between the means.

```{r}
abs(mean(numericStepsPerDay[ , 2]) - mean(stepsPerDay[ , 2]))
```

And between the medians.
```{r}
abs(median(numericStepsPerDay[ , 2]) - median(stepsPerDay[ , 2]))
```

This result should make it obvious that ignoring NA values does affect the
results of calculations made using those values. It should also be apparent that 
the strategy used to replace the NA values will impact the results, so it should
be chosen with care.

## Are there differences in activity patterns between weekdays and weekends?

**TODO** *This part of the report needs to be completed.*